---
title: "Rearrangement and selection of epitope-specific T-cell clonotypes"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table)
library(dplyr)
library(ggplot2)
library(reshape2)
library(scales)
library(stringr)

select = dplyr::select
```

### Rearrangement model suppl

Pre-processing and selecting epitopes

```{r}
dt.vdjdb = fread("rearr_model/VDJDB_fullP_rob_ageing.txt") 
```

Fraction of non-zero Prearr TCRs

```{r}
dt.epicount1 = dt.vdjdb %>%
  group_by(antigen.epitope) %>%
  summarise(total = n(), 
            total_r_a = sum(genP_0mism_ageing > 0),
            frac_r_a = sum(genP_0mism_ageing > 0) / n(),
            log_mean_p_a = sum(ifelse(genP_0mism_ageing > 0, log10(genP_0mism_ageing), 0)) / sum(genP_0mism_ageing > 0),
            total_r_h = sum(genP_0mism_rob > 0),
            frac_r_h = sum(genP_0mism_rob > 0) / n(),
            log_mean_p_h = sum(ifelse(genP_0mism_rob > 0, log10(genP_0mism_rob), 0)) / sum(genP_0mism_rob > 0))

ggplot(dt.epicount1) +
  geom_histogram(aes(pmin(30,total)), binwidth = 1, color= "black", fill=NA) +
  geom_histogram(aes(pmin(30,total_r_a)), binwidth = 1, color = "red", fill=NA) +
  geom_histogram(aes(pmin(30,total_r_h)), binwidth = 1, color = "blue", fill=NA) +
  ylab("Number of epitopes") + xlab("Number of specific TCRs with non-zero Prearr") +
  theme_bw()

ggplot(dt.epicount1 %>% filter(total > 30)) +
  geom_point(aes(x = frac_r_a, y = log_mean_p_a), color = "red") +
  geom_smooth(aes(x = frac_r_a, y = log_mean_p_a), method = "lm", color = "red") +
  geom_point(aes(x = frac_r_h, y = log_mean_p_h), color = "blue") +
  geom_smooth(aes(x = frac_r_h, y = log_mean_p_h), method = "lm", color = "blue") +
  ylab("Mean log Prearr for non-zero Prearr") + xlab("Fraction of non-zero Prearr") +
  theme_bw()

print(with(dt.epicount1 %>% filter(total > 30), cor.test(frac_r_a, log_mean_p_a)))
print(with(dt.epicount1 %>% filter(total > 30), cor.test(frac_r_h, log_mean_p_h)))
```

Fraction of non-zero Prearr TCRs

```{r}
dt.epicount2 = dt.vdjdb %>%
  group_by(antigen.epitope) %>%
  summarise(total = n(), 
            total_r_a = sum(genP_1mism_ageing > 0),
            frac_r_a = sum(genP_1mism_ageing > 0) / n(),
            log_mean_p_a = sum(ifelse(genP_1mism_ageing > 0, log10(genP_1mism_ageing), 0)) / sum(genP_1mism_ageing > 0),
            total_r_h = sum(genP_1mism_rob > 0),
            frac_r_h = sum(genP_1mism_rob > 0) / n(),
            log_mean_p_h = sum(ifelse(genP_1mism_rob > 0, log10(genP_1mism_rob), 0)) / sum(genP_1mism_rob > 0))

ggplot(dt.epicount2) +
  geom_histogram(aes(pmin(30,total)), binwidth = 1, color= "black", fill=NA) +
  geom_histogram(aes(pmin(30,total_r_a)), binwidth = 1, color = "red", fill=NA) +
  geom_histogram(aes(pmin(30,total_r_h)), binwidth = 1, color = "blue", fill=NA) +
  ylab("Number of epitopes") + xlab("Number of specific TCRs with non-zero Prearr/1mism") +
  theme_bw() 

ggplot(dt.epicount2 %>% filter(total > 30)) +
  geom_point(aes(x = frac_r_a, y = log_mean_p_a), color = "red") +
  geom_point(aes(x = frac_r_h, y = log_mean_p_h), color = "blue") +
  ylab("Mean log Prearr/1mism for non-zero Prearr/1mism") + xlab("Fraction of non-zero Prearr/1mism") +
  theme_bw()

print(with(dt.epicount2 %>% filter(total > 30), cor.test(frac_r_a, log_mean_p_a)))
print(with(dt.epicount2 %>% filter(total > 30), cor.test(frac_r_h, log_mean_p_h)))
```

Correlation between rearrangement probs from two models, with and without mismatches

```{r}
ggplot(dt.vdjdb, aes(x=genP_0mism_ageing,y=genP_1mism_ageing)) +
  geom_point(shape=21) +
  geom_smooth(method = "lm") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  scale_x_log10(limits=c(1e-13,1e-4)) +
  scale_y_log10(limits=c(1e-13,1e-4)) +
  theme_bw()

ggplot(dt.vdjdb, aes(x=genP_0mism_rob,y=genP_1mism_rob)) +
  geom_point(shape=21) +
  geom_smooth(method = "lm") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  scale_x_log10(limits=c(1e-13,1e-4)) +
  scale_y_log10(limits=c(1e-13,1e-4)) +
  theme_bw()

ggplot(dt.vdjdb, aes(x=genP_0mism_ageing,y=genP_0mism_rob)) +
  geom_point(shape=21) +
  geom_smooth(method = "lm") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  scale_x_log10(limits=c(1e-13,1e-4)) +
  scale_y_log10(limits=c(1e-13,1e-4)) +
  theme_bw()

ggplot(dt.vdjdb, aes(x=genP_1mism_ageing,y=genP_1mism_rob)) +
  geom_point(shape=21) +
  geom_smooth(method = "lm") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  scale_x_log10(limits=c(1e-13,1e-4)) +
  scale_y_log10(limits=c(1e-13,1e-4)) +
  theme_bw()
```

### Annotated data suppl

```{r}
df.hip.raw = fread("zcat annotations/hip_annot_1.txt.gz")

HIP_R = sum(fread("annotations/hip_stats.txt")$diversity)
HIP_S = sum(nrow(fread("annotations/hip_stats.txt")))

df.hip = df.hip.raw %>%
  group_by(cdr3, antigen.epitope, antigen.species, mhc.a, mhc.b, mhc.class) %>%
  summarise(count = sum(occurrences), 
            freq = sum(occurrences) / HIP_R,
            incidence = sum(incidence) / HIP_S / n(),
            convergence = sum(convergence) / sum(occurrences)) %>%
  merge(dt.vdjdb %>% select(cdr3, antigen.epitope, antigen.species, mhc.a, mhc.b, mhc.class, 
                            genP_1mism_rob),
        by = c("cdr3", "antigen.epitope", 'antigen.species', "mhc.a", 'mhc.b', 'mhc.class'), all.y = T) %>%
  mutate(count = ifelse(is.na(count), 0, count),
         freq = ifelse(is.na(freq), 0, freq),
         incidence = ifelse(is.na(incidence), 0, incidence),
         convergence = ifelse(is.na(convergence), 0, convergence))
```

```{r}
fancy_scientific = function(l) {
     # turn in to character string in scientific notation
     l = format(l, scientific = TRUE)
     # quote the part before the exponent to keep all the digits
     l = gsub("^(.*)e", "'\\1'e", l)
     # turn the 'e+' into plotmath format
     l = gsub("e", "%*%10^", l)
     # return this as an expression
     parse(text=l)
}

lm_eqn = function(ll){
    eq = substitute(italic(logY) == a + b %.% italic(logX)*","~~italic(R)^2~"="~r2, 
         list(a = format(coef(ll)[1], digits = 2), 
              b = format(coef(ll)[2], digits = 2), 
             r2 = format(summary(ll)$r.squared, digits = 2)))
    as.character(as.expression(eq))
}

rr = with(df.hip %>% filter(genP_1mism_rob > 0, freq > 0), cor.test(log10(genP_1mism_rob), log10(freq)))
ll = with(df.hip %>% filter(genP_1mism_rob > 0, freq > 0), lm(log10(freq) ~ log10(genP_1mism_rob)))

p1=ggplot(df.hip, aes(x=genP_1mism_rob,y=freq)) +
  stat_density_2d(geom = "hex", aes(fill=..density..), contour = F) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +
  annotate(geom = "label", x = 1e-7, y = 1e-12, label = lm_eqn(ll), parse = T) +
  scale_x_log10("Theoretical rearrangement probability", limits=c(1e-13,1e-3), labels = fancy_scientific) +
  scale_y_log10("Observed rearrangement frequency", limits=c(1e-13,1e-3), labels = fancy_scientific) +
  scale_fill_gradient("Density", low = "white", high = "black") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        aspect.ratio = 1, legend.position = c(0.15, 0.75))
p1
```

```{r}
length(unique((df.hip %>% filter(genP_1mism_rob > 0, freq > 0))$cdr3))
print(HIP_R)
print(HIP_S)
```

Expected and observed difference between occurrences

```{r}
obs_ratio = as.numeric(10^(-coef(ll)[1]))

mean_stop_prob = (df.hip.raw %>%
  filter(DStart > 0) %>%
  select(cdr3nt, VEnd, DStart, DEnd, JStart) %>%
  unique %>%
  summarise(mean_stop_prob = mean((61/64)^round((DStart - VEnd + JStart - DEnd + 2)/3))))$mean_stop_prob

coding_prob = (1-2/3) * mean_stop_prob

obs_ratio
coding_prob
```

```{r}
ggplot(df.hip, aes(x=genP_1mism_rob*3,y=incidence)) +
  geom_point(shape=21) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_discrete(guide=F) +
  theme_bw()
```

```{r}
ggplot(df.hip %>% filter(incidence > 0), aes(x=incidence, y=convergence)) +
  geom_point(shape=21) +
 # geom_smooth() +
  geom_vline(xintercept = 0.05, color = "red", linetype = "dashed") +
  scale_color_discrete(guide=F) +
  theme_bw()

ggplot(df.hip %>% filter(incidence > 0), aes(x=genP_1mism_rob, y=convergence)) +
  geom_point(shape=21) +
 # geom_smooth() +
  geom_vline(xintercept = 1e-6, color = "red", linetype = "dashed") +
  scale_color_discrete(guide=F) +
  theme_bw()
```

```{r}
p2=ggplot(df.hip %>% filter(convergence > 0), aes(x=genP_1mism_rob,y=convergence)) +
  stat_density_2d(geom = "tile", aes(fill=..density..), contour = F) +
  geom_smooth(color = "red") +
  scale_x_log10("Theoretical rearrangement probability", limits=c(1e-13,1e-3), labels = fancy_scientific) +
  scale_y_continuous("Unique variants per rearrangement (convergence)", expand = c(0,0), limits = c(-0.0001, 1.0001)) +
  scale_fill_gradient("Density", low = "white", high = "black") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        aspect.ratio = 1, legend.position = c(0.15, 0.25))
p2
```

```{r}
dt.epicount3 = df.hip %>%
  group_by(antigen.epitope) %>%
  summarise(total = n(), 
            found_frac = sum(count > 0) / total, 
            found_p_frac = sum(genP_1mism_rob > 0) / total)

ggplot(dt.epicount3, aes(x = total, y = found_frac)) +
  geom_point(shape = 21, aes(color = total > 30 & found_frac > 0.75)) +
  theme_bw()

ggplot(dt.epicount3, aes(x = total, y = found_p_frac)) +
  geom_point(shape = 21, aes(color = total > 30 & found_frac > 0.75)) +
  theme_bw()

good_epi = (dt.epicount3 %>% filter(total > 30 & found_frac > 0.75))$antigen.epitope

df.hip.all = df.hip
df.hip = df.hip %>% filter(antigen.epitope %in% good_epi, mhc.class == "MHCI")
```


## Exploratory analysis

### By epitope

```{r}
df.hip.gp = df.hip %>%
  filter(genP_1mism_rob > 0)
df.hip.gp.s = df.hip.gp %>%
  group_by(antigen.epitope) %>%
  summarise(genP_med = median(genP_1mism_rob)) %>%
  arrange(-genP_med)
df.hip.gp$antigen.epitope = factor(df.hip.gp$antigen.epitope, 
                                   levels = df.hip.gp.s$antigen.epitope)

ggplot(df.hip.gp, aes(x=antigen.epitope, group = antigen.epitope, y=genP_1mism_rob, fill = antigen.species)) +
  geom_violin() + stat_summary(fun.y=median, geom="point", shape=21, fill = "white", color="black") +
  scale_y_log10(breaks = 10^(-15:-1),
                label= function(x) {ifelse(x==0, "0", parse(text=gsub("[+]", "", gsub(".+e", "10^", scientific_format()(x)))))} ) +
  coord_flip() +
  scale_fill_brewer(palette = "Set3") +
  theme_bw()

a = kruskal.test(freq ~ antigen.epitope, data = df.hip.gp)
print(a)
```

```{r}
p3=ggplot(df.hip.gp, aes(x=antigen.epitope, group = antigen.epitope, y=genP_1mism_rob)) +
  geom_violin(fill = "black") + 
  stat_summary(fun.y=median, geom="point", color="red") +
  scale_y_log10("Theoretical rearrangement probability",
                breaks = 10^(-15:-1),
                label= function(x) {ifelse(x==0, "0", parse(text=gsub("[+]", "", gsub(".+e", "10^", scientific_format()(x)))))} ) +
  xlab("") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid.major.y = element_line(colour="grey50", linetype="dashed"),
        panel.grid.minor.y = element_line(colour="grey50", linetype="dotted"),
        panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank())
p3

length(unique(df.hip.gp$antigen.epitope))

a = kruskal.test(genP_1mism_rob ~ I(as.factor(antigen.epitope)), data = df.hip.gp)
print(a)
print(a$p.value)
a = aov(log(genP_1mism_rob) ~ I(as.factor(antigen.epitope)), data = df.hip.gp)
print(summary(a)[[1]][["Pr(>F)"]][1])
summary(a)
```

```{r}
tmp = df.hip.gp %>%
  group_by(antigen.epitope) %>%
  summarise(genP = median(genP_1mism_rob), count = n())

p26= ggplot(tmp, aes(genP, count)) +
  geom_smooth(method = "lm", color = "red") +
  geom_point() +
  ylab("Number of unique VDJdb TCRs") +
  scale_x_log10("Theoretical rearrangement probability", limits = c(1e-8,1e-6)) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        aspect.ratio = 1, legend.position = c(0.15, 0.25))
p26

print(cor.test(log(tmp$genP), tmp$count))
```


```{r}
print(length(unique(df.hip.gp$antigen.epitope)))
```

```{r}
df.hip.obs = df.hip %>%
  filter(freq > 0)
df.hip.obs.s = df.hip.obs %>%
  group_by(antigen.epitope) %>%
  summarise(freq_med = median(freq)) %>%
  arrange(-freq_med)
df.hip.obs$antigen.epitope = factor(df.hip.obs$antigen.epitope, 
                                   levels = df.hip.obs.s$antigen.epitope)

ggplot(df.hip.obs, aes(x=antigen.epitope, group = antigen.epitope, y=freq, fill = antigen.species)) +
  geom_violin() + stat_summary(fun.y=median, geom="point", shape=21, fill = "white", color="black") +
  scale_y_log10(breaks = 10^(-15:-1),
                label= function(x) {ifelse(x==0, "0", parse(text=gsub("[+]", "", gsub(".+e", "10^", scientific_format()(x)))))} ) +
  coord_flip() +
  scale_fill_brewer(palette = "Set3") +
  theme_bw()

a = kruskal.test(freq ~ antigen.epitope, data = df.hip.obs)
print(a)
```

```{r}
df.hip.odds = df.hip %>%
  filter(freq > 0, genP_1mism_rob > 0) %>%
  mutate(odds = freq/genP_1mism_rob,
         log.odds = log(odds))
df.hip.odds.s = df.hip.odds %>%
  group_by(antigen.epitope) %>%
  summarise(odds_med = median(odds)) %>%
  arrange(-odds_med)
df.hip.odds$antigen.epitope = factor(df.hip.odds$antigen.epitope, 
                                   levels = df.hip.odds.s$antigen.epitope)

ggplot(df.hip.odds, aes(x=antigen.epitope, group = antigen.epitope, y=odds/3, fill = antigen.species)) +
  geom_violin() + stat_summary(fun.y=median, geom="point", shape=21, fill = "white", color="black") +
  scale_y_log10(breaks = 2^(seq(-2,20,by=2))) +
  coord_flip() +
  scale_fill_brewer(palette = "Set3") +
  theme_bw()

a = kruskal.test(odds ~ antigen.epitope, data = df.hip.odds)
print(a)
```

### By species

```{r}
df.hip.gp = df.hip %>%
  filter(genP_1mism_rob > 0)
df.hip.gp.s = df.hip.gp %>%
  group_by(antigen.species) %>%
  summarise(genP_med = median(genP_1mism_rob)) %>%
  arrange(-genP_med)
df.hip.gp$antigen.species = factor(df.hip.gp$antigen.species, 
                                   levels = df.hip.gp.s$antigen.species)

ggplot(df.hip.gp, aes(x=antigen.species, group = antigen.species, y=genP_1mism_rob)) +
  geom_violin() + stat_summary(fun.y=median, geom="point", shape=21, fill = "white", color="black") +
  scale_y_log10(breaks = 10^(-15:-1),
                label= function(x) {ifelse(x==0, "0", parse(text=gsub("[+]", "", gsub(".+e", "10^", scientific_format()(x)))))} ) +
  coord_flip() +
  theme_bw()

kk = kruskal.test(freq ~ antigen.species, data = df.hip.gp)
print(kk)

aa = aov(log(genP_1mism_rob) ~ antigen.species, data = df.hip.gp)
summary(aa)
TukeyHSD(aa, "antigen.species")
```

```{r}
p4=ggplot(df.hip.gp, aes(x=antigen.species, y=genP_1mism_rob)) +
  geom_violin(aes(group = antigen.species)) + 
  geom_violin(fill = "black") + 
  stat_summary(fun.y=median, geom="point", color="red") +
  scale_y_log10("Theoretical rearrangement probability",
                breaks = 10^(-15:-1),
                label= function(x) {ifelse(x==0, "0", parse(text=gsub("[+]", "", gsub(".+e", "10^", scientific_format()(x)))))} ) +
  xlab("") +
  theme_bw() +
  theme(aspect.ratio = 1, axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid.major.y = element_line(colour="grey50", linetype="dashed"),
        panel.grid.minor.y = element_line(colour="grey50", linetype="dotted"),
        panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank())
p4

a = kruskal.test(genP_1mism_rob ~ I(as.factor(antigen.species)), data = df.hip.gp)
print(a)
print(a$p.value)
a = aov(log(genP_1mism_rob) ~ I(as.factor(antigen.species)), data = df.hip.gp)
print(summary(a)[[1]][["Pr(>F)"]][1])
summary(a)
```

```{r}
df.hip.obs = df.hip %>%
  filter(freq > 0)
df.hip.obs.s = df.hip.obs %>%
  group_by(antigen.species) %>%
  summarise(freq_med = median(freq)) %>%
  arrange(-freq_med)
df.hip.obs$antigen.species = factor(df.hip.obs$antigen.species, 
                                   levels = df.hip.obs.s$antigen.species)

ggplot(df.hip.obs, aes(x=antigen.species, group = antigen.species, y=freq)) +
  geom_violin() + stat_summary(fun.y=median, geom="point", shape=21, fill = "white", color="black") +
  scale_y_log10(breaks = 10^(-15:-1),
                label= function(x) {ifelse(x==0, "0", parse(text=gsub("[+]", "", gsub(".+e", "10^", scientific_format()(x)))))} ) +
  coord_flip() +
  theme_bw()

kk = kruskal.test(freq ~ antigen.species, data = df.hip.obs)
print(kk)

aa = aov(log(freq) ~ antigen.species, data = df.hip.obs)
summary(aa)
TukeyHSD(aa, "antigen.species")
```

```{r}
df.hip.odds = df.hip %>%
  filter(freq > 0, genP_1mism_rob > 0) %>%
  mutate(odds = freq/genP_1mism_rob,
         log.odds = log(odds))
df.hip.odds.s = df.hip.odds %>%
  group_by(antigen.species) %>%
  summarise(odds_med = median(odds)) %>%
  arrange(-odds_med)
df.hip.odds$antigen.species = factor(df.hip.odds$antigen.species, 
                                   levels = df.hip.odds.s$antigen.species)

ggplot(df.hip.odds, aes(x=antigen.species, group = antigen.species, y=odds/3)) +
  geom_violin() + stat_summary(fun.y=median, geom="point", shape=21, fill = "white", color="black") +
  scale_y_log10(breaks = 2^(seq(-2,20,by=2))) +
  coord_flip() +
  theme_bw()

kk = kruskal.test(odds ~ antigen.species, data = df.hip.odds)
print(kk)

aa = aov(log(odds) ~ antigen.species, data = df.hip.odds)
summary(aa)
TukeyHSD(aa, "antigen.species")
```

### Number of publics

```{r}
df.hip.frac = df.hip %>%
  group_by(antigen.epitope) %>%
  summarise(publ_5 = sum(incidence > 0.05) / n(),
            publ_10 = sum(incidence > 0.1) / n(),
            publ_15 = sum(incidence > 0.15) / n(),
            publ_20 = sum(incidence > 0.20) / n())

df.hip.frac$antigen.epitope = with(df.hip.frac,
                                   factor(antigen.epitope, levels = antigen.epitope[order(-publ_5)]))

df.hip.frac = df.hip.frac %>%
  melt

df.hip.frac$publ = paste0(str_split_fixed(df.hip.frac$variable, "_", n = 2)[,2],"%")
tmp = unique(df.hip.frac$publ)
df.hip.frac$publ = factor(df.hip.frac$publ, levels = tmp[order(sapply(tmp, function(x) -as.numeric(gsub("%", "", x))))])

length(unique(df.hip.frac$antigen.epitope))

p5=ggplot(df.hip.frac, 
       aes(x=antigen.epitope, y = value, fill = publ)) +
  geom_bar(stat="identity", color = "black", position = "identity") +
  scale_fill_brewer("Publicity", palette = "RdBu") +
  scale_x_discrete("") + 
  scale_y_continuous("Fraction of clonotypes", expand = c(0,0), limits = c(0,0.35)) +
  theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p5
```

```{r}
df.hip.frac = df.hip %>%
  group_by(antigen.species) %>%
  summarise(publ_5 = sum(incidence > 0.05) / n(),
            publ_10 = sum(incidence > 0.1) / n(),
            publ_15 = sum(incidence > 0.15) / n(),
            publ_20 = sum(incidence > 0.20) / n())

df.hip.frac$antigen.species = with(df.hip.frac,
                                   factor(antigen.species, levels = antigen.species[order(-publ_5)]))

df.hip.frac = df.hip.frac %>%
  melt

df.hip.frac$publ = paste0(str_split_fixed(df.hip.frac$variable, "_", n = 2)[,2],"%")
tmp = unique(df.hip.frac$publ)
df.hip.frac$publ = factor(df.hip.frac$publ, levels = tmp[order(sapply(tmp, function(x) -as.numeric(gsub("%", "", x))))])

p6=ggplot(df.hip.frac, 
       aes(x=antigen.species, y = value, fill = publ)) +
  geom_bar(stat="identity", color = "black", position = "identity") +
  scale_fill_brewer("Publicity", palette = "RdBu", guide = F) +
  scale_x_discrete("") + 
  scale_y_continuous("Fraction of clonotypes", expand = c(0,0), limits = c(0,0.35)) +
  theme_bw() +
    theme(aspect.ratio = 1, axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6
```

## Epitope properties and rearrangement prob/selection

```{r}
df.hip.all.1 = df.hip.all %>% filter(mhc.class == "MHCI", nchar(antigen.epitope) %in% c(8:11))
```

```{r}
library(Hmisc)
dt.epi.prop = rbindlist(lapply(strsplit(unique(df.hip.all.1$antigen.epitope), split = ""),
                             function(x) data.table(aa = x,
                                                    antigen.epitope = paste0(x, collapse = "")))) 

dt.epi.prop = rbind(dt.epi.prop %>% merge(fread("kidera.txt") %>% mutate(len = 1) %>% melt, by = "aa", allow.cartesian = T),
                    dt.epi.prop %>% merge(fread("martin-lavery.txt") %>% melt, by = "aa", allow.cartesian = T)) %>%
  group_by(antigen.epitope, variable) %>%
  summarise(value = sum(value)) %>%
  group_by(variable) %>%
  mutate(value_c = as.factor(ifelse(variable == "len", as.integer(as.factor(value)), as.integer(cut2(value, g=4)))))
```

```{r}
ggplot(dt.epi.prop, aes(x = antigen.epitope, y = value, color = value_c)) +
  geom_point() +
  facet_wrap(~variable, scales = "free") +
  scale_color_brewer("Factor level", palette = "RdBu") +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

```{r}
dt.hip.prop = dt.epi.prop %>% merge(df.hip.all.1, by = "antigen.epitope", allow.cartesian = T)
```

```{r}
ggplot(dt.hip.prop, aes(x = value_c, y = genP_1mism_rob)) +
  geom_violin(aes(group = value_c, fill = value_c), draw_quantiles = c(0.25,0.5,0.75)) +
  #stat_summary(fun.y=mean, geom="point", shape=21, fill = "white", color="black") +
  scale_y_log10() +
  scale_fill_brewer(guide = F, palette = "RdBu") +
  facet_wrap(~variable) +
  theme_bw()

ggplot(dt.hip.prop, aes(x = value_c, group = value_c, y = freq, fill = value_c)) +
  geom_violin() +
  stat_summary(fun.y=median, geom="point", shape=21, fill = "white", color="black") +
  scale_y_log10() +
  scale_fill_brewer(guide = F, palette = "RdBu") +
  facet_wrap(~variable) +
  theme_bw()
```

```{r}
dt.hip.prop.cor = dt.hip.prop %>%
  group_by(variable) %>%
  summarise(r_freq = cor.test(freq, value, method = "spearman")$estimate,
            p_freq = cor.test(freq, value, method = "spearman")$p.value,
            r_P = cor.test(genP_1mism_rob, value, method = "spearman")$estimate,
            p_P = cor.test(genP_1mism_rob, value, method = "spearman")$p.value)
```

```{r}
dt.hip.prop.aov1 = dt.hip.prop %>%
  filter(freq > 0) %>%
  group_by(variable) %>%
  summarise(F_freq = summary(aov(log(freq1)~value_c1, data.table(freq1 = freq, value_c1 = as.integer(value_c))))[[1]][["F value"]][1],
            p_freq = summary(aov(log(freq1)~value_c1, data.table(freq1 = freq, value_c1 = as.integer(value_c))))[[1]][["Pr(>F)"]][1]) %>%
  mutate(p_freq_adj = p.adjust(p_freq))

dt.hip.prop.aov2 = dt.hip.prop %>%
  filter(genP_1mism_rob > 0) %>%
  group_by(variable) %>%
  summarise(F_P = summary(aov(log(P1)~value_c1, data.table(P1 = genP_1mism_rob, value_c1 = as.integer(value_c))))[[1]][["F value"]][1],
            pa_P = summary(aov(log(P1)~value_c1, data.table(P1 = genP_1mism_rob, value_c1 = as.integer(value_c))))[[1]][["Pr(>F)"]][1]) %>%
  mutate(pa_P_adj = p.adjust(pa_P))
```

```{r}
dt.hip.prop.cor1 = dt.hip.prop %>%
  filter(freq > 0) %>%
  group_by(variable) %>%
  summarise(r_freq = cor.test(freq, value, method = "spearman")$estimate,
            p_freq = cor.test(freq, value, method = "spearman")$p.value) %>%
  mutate(p_freq_adj = p.adjust(p_freq, method = "BH"))

dt.hip.prop.cor2 = dt.hip.prop %>%
  filter(genP_1mism_rob > 0) %>%
  group_by(variable) %>%
  summarise(r_P = cor.test(genP_1mism_rob, value, method = "spearman")$estimate,
            pr_P = cor.test(genP_1mism_rob, value, method = "spearman")$p.value) %>%
  mutate(pr_P_adj = p.adjust(pr_P))
```

```{r}
dt.hip.prop.1 = merge(dt.hip.prop.aov2, dt.hip.prop.cor2)
print(dt.hip.prop.1)

print(dt.hip.prop.1 %>% filter(pr_P_adj < 0.01 & pa_P_adj < 0.01))
```

```{r}
df.hip.frac.prop = df.hip %>%
  merge(dt.hip.prop %>% 
          select(antigen.epitope, variable, value_c) %>% 
          unique %>%
          merge(dt.hip.prop.1 %>% filter(pr_P_adj < 0.01 & pa_P_adj < 0.01) %>% select(variable) %>% unique), 
        by = "antigen.epitope", allow.cartesian = T) %>%
  group_by(variable, value_c) %>%
  summarise(publ_5 = sum(incidence > 0.05) / n(),
            publ_10 = sum(incidence > 0.1) / n(),
            publ_15 = sum(incidence > 0.15) / n(),
            publ_20 = sum(incidence > 0.20) / n(),
            tmp = n())


df.hip.frac.prop$property = df.hip.frac.prop$variable
df.hip.frac.prop$variable = NULL
df.hip.frac.prop$property_level = df.hip.frac.prop$value_c
df.hip.frac.prop$value_c = NULL

df.hip.frac.prop = df.hip.frac.prop %>%
  melt

df.hip.frac.prop$variable = with(df.hip.frac.prop,
                                 factor(paste0(str_split_fixed(variable, "_", 2)[,2], "%"), levels = rev(c("5%", "10%", "15%", "20%"))))

df.hip.frac.prop$property_level2 = factor(with(df.hip.frac.prop,
                                         ifelse(property == "len", as.character(as.integer(property_level) + 7), paste0("Q",property_level))), 
                                         levels = c("Q1","Q2","Q3","Q4", "8", "9", "10", "11"))
df.hip.frac.prop$property2 = with(df.hip.frac.prop,
                                  ifelse(property=="len","Length", ifelse(property=="f6", "Partial specific volume", "Surrounding hydrophobicity")))
# 6 - Partial specific volume,
# 10 - Surrounding hydrophobicity
p7=ggplot(df.hip.frac.prop, 
       aes(x=property_level2, y = value, 
           fill = variable)) +
  geom_bar(stat="identity", color = "black", position = "identity") +
  scale_fill_brewer("Population frequency", palette = "RdBu") +
  scale_x_discrete("") + 
  scale_y_continuous("Fraction of clonotypes", expand = c(0,0), limits = c(0,0.25)) +
  facet_wrap(~property2, scales = "free_x") +
  theme_bw() +
    theme(legend.position = "bottom", axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        strip.background = element_blank())
p7
```

## Figures

```{r}
ggsave("figures/p1.pdf", p1, width = 4, height = 4)
ggsave("figures/p2.pdf", p2, width = 4, height = 4)
ggsave("figures/p3.pdf", p3, width = 4*2, height = 4)
ggsave("figures/p4.pdf", p4, width = 4, height = 4)
ggsave("figures/p5.pdf", p5, width = 4*2, height = 4)
ggsave("figures/p6.pdf", p6, width = 4, height = 4)
ggsave("figures/p7.pdf", p7, width = 4*2, height = 4)
ggsave("figures/p26.pdf", p26, width = 4, height = 4)
```





