---
title: "AB pairing"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table)
library(dplyr)
library(reshape2)
library(ggplot2)
library(scales)
library(forcats)
library(gridExtra)
```

### Preprocessing pairseq data

PairSEQ data by Howie

```{r}
dt.pairs = fread("zcat tcr_ab_pairs.txt.gz") %>%
  filter(startsWith(aaSeqCDR3, "C"), endsWith(aaSeqCDR3, "F") | endsWith(aaSeqCDR3, "W"))
```

Transform table, count the number of rearrangements

```{r}
dt.pairs.1 = dt.pairs %>%
  filter(chain == "A") %>%
  mutate(nSeqCDR3.A = nSeqCDR3, 
         aaSeqCDR3.A = aaSeqCDR3, 
         allVGenes.A = allVGenes,
         allJGenes.A = allJGenes) %>%
  select(sample, clone, nSeqCDR3.A, aaSeqCDR3.A, allVGenes.A, allJGenes.A) %>%
  merge(dt.pairs %>%
          filter(chain == "B") %>%
          mutate(nSeqCDR3.B = nSeqCDR3, 
                 aaSeqCDR3.B = aaSeqCDR3, 
                 allVGenes.B = allVGenes, 
                 allJGenes.B = allJGenes) %>%
          select(sample, clone, nSeqCDR3.B, aaSeqCDR3.B, allVGenes.B, allJGenes.B)) %>%
  select(-clone) %>%
  unique %>%
  group_by(aaSeqCDR3.A, aaSeqCDR3.B) %>%
  summarise(count = n(), 
            allVGenes.A = allVGenes.A[1], allVGenes.B = allVGenes.B[1],
            allJGenes.A = allJGenes.A[1], allJGenes.B = allJGenes.B[1],
            nSeqCDR3.A = nSeqCDR3.A[1], nSeqCDR3.B = nSeqCDR3.B[1])
```

### Annotation

Create mock VDJtools format datasets

```{r}
dt.pairs.1 %>%
  group_by(CDR3aa = aaSeqCDR3.A) %>%
  summarise(count = n(), CDR3nt = nSeqCDR3.A[1], 
            V = allVGenes.A[1], D = "", J = allJGenes.A[1]) %>%
  ungroup %>%
  mutate(freq = count / sum(count)) %>%
  select(count, freq, CDR3nt, CDR3aa, V, D, J) %>%
  fwrite("pairs.a.txt", sep = "\t")

dt.pairs.1 %>%
  group_by(CDR3aa = aaSeqCDR3.B) %>%
  summarise(count = n(), CDR3nt = nSeqCDR3.B[1], 
            V = allVGenes.B[1], D = "", J = allJGenes.B[1]) %>%
  ungroup %>%
  mutate(freq = count / sum(count)) %>%
  select(count, freq, CDR3nt, CDR3aa, V, D, J) %>%
  fwrite("pairs.b.txt", sep = "\t")
```

Annotate

```{r}
system("rm pairs.a.txt.gz")
system("gzip pairs.a.txt")
system("rm pairs.b.txt.gz")
system("gzip pairs.b.txt")
system("java -Xmx40G -jar ../vdjdb_new/vdjdb-1.1.6.jar -S human -R TRA --search-scope 1,0,0,1 pairs.a.txt.gz a")
system("java -Xmx40G -jar ../vdjdb_new/vdjdb-1.1.6.jar -S human -R TRB --search-scope 1,0,0,1 pairs.b.txt.gz b")
```

### Annotation statistics

Select epitopes that have both alpha and beta

```{r}
dt.epi = fread("../vdjdb_new/vdjdb.slim.txt") %>%
  filter(species == "HomoSapiens") %>%
  group_by(gene, antigen.epitope) %>%
  summarise(db.count = length(unique(cdr3))) %>%
  filter(db.count >= 30) %>%
  group_by(antigen.epitope) %>%
  mutate(chains = length(unique(gene))) %>%
  filter(chains == 2) %>%
  select(-chains) %>%
  filter(!startsWith(antigen.epitope, "FRD") &
           !startsWith(antigen.epitope, "LLL")) # unpaired only
```

Load annotated sequences

```{r}
dt.ann.a = fread("a.pairs.a.annot.txt") %>%
  mutate(cdr3.a = cdr3aa) %>%
  select(count, cdr3.a, antigen.epitope) %>%
  unique %>%
  merge(dt.epi %>% 
          filter(gene == "TRA") %>%
          mutate(db.count.a = db.count) %>%
          select(db.count.a, antigen.epitope))

dt.ann.b = fread("b.pairs.b.annot.txt") %>%
  filter(antigen.epitope %in% dt.epi$antigen.epitope) %>%
  mutate(cdr3.b = cdr3aa) %>%
  select(count, cdr3.b, antigen.epitope) %>%
  unique %>%
  merge(dt.epi %>% 
          filter(gene == "TRB") %>%
          mutate(db.count.b = db.count) %>%
          select(db.count.b, antigen.epitope))
```

Paired annotations

```{r}
dt.ann.ab = dt.pairs.1 %>%
  ungroup %>%
  mutate(cdr3.a = aaSeqCDR3.A, cdr3.b = aaSeqCDR3.B) %>%
  select(count, cdr3.a, cdr3.b) %>%
  merge(dt.ann.a %>% select(-count)) %>%
  merge(dt.ann.b %>% select(-count))
```

Plot annotated percent

```{r fig.width=4, fig.height=4}
total = sum(dt.pairs.1$count)
pa = dt.ann.a %>% .$count %>% sum / total
pb = dt.ann.b %>% .$count %>% sum / total
pab = dt.ann.ab %>% .$count %>% sum / total

ann_stats = data.frame(type = factor(c("A", "B", "AB"), 
                                     levels = c("A", "B", "AB")),
                       p = c(pa, pb, pab))

p1 = ggplot(ann_stats, aes(x = type, y = p, fill = type)) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous("Matched to VDJdb", labels = percent) +
  scale_fill_brewer(guide = F, palette = "Set1") +
  xlab("") +
  theme_bw() +
  theme(aspect = 1, panel.grid.major = element_blank(), panel.grid.minor = element_blank())

p1

pdf("p1.pdf", width = 4, height = 4)
p1
dev.off()
```

### Statistics by epitope

Alpha versus beta

```{r fig.width=4, fig.height=4}
dt.ann.a.s = dt.ann.a %>%
  group_by(antigen.epitope, db.count.a) %>%
  summarise(count.a = sum(count)) %>%
  merge(dt.epi %>% 
          filter(gene == "TRA") %>%
          mutate(db.count.a = db.count) %>%
          select(db.count.a, antigen.epitope), all.y = T) %>%
  mutate(count.a = ifelse(is.na(count.a), 0, count.a),
         pa = count.a / total / db.count.a)

dt.ann.b.s = dt.ann.b %>%
  group_by(antigen.epitope, db.count.b) %>%
  summarise(count.b = sum(count)) %>%
  merge(dt.epi %>% 
          filter(gene == "TRB") %>%
          mutate(db.count.b = db.count) %>%
          select(db.count.b, antigen.epitope), all.y = T) %>%
  mutate(count.b = ifelse(is.na(count.b), 0, count.b),
         pb = count.b / total / db.count.b)

dt.ann.ab.s1 = dt.ann.a.s %>%
  select(pa, antigen.epitope) %>%
  merge(dt.ann.b.s %>%
          select(pb, antigen.epitope)) %>%
  filter(pa > 0 & pb > 0)

cres = with(dt.ann.ab.s1, cor.test(pa, pb))
lbl = paste0("R=", round(cres$estimate, 2), ", P=", round(cres$p.value, 2))

p2=ggplot(dt.ann.ab.s1, aes(x = pa, y = pb)) +
  annotate("text", x = 3.5e-5, y = 2.3e-5, label = lbl) +
  geom_smooth(method = "lm") +
  geom_text(aes(y = pb + 1e-6, label = substr(antigen.epitope, 1, 3))) +
  geom_point(size = 3) +
  xlab("Mean TRA frequency") +
  ylab("Mean TRB frequency") +
  #annotation_custom(grob = tt, xmin = 5e-5, xmax = 7e-5, ymin = 2.5e-6, ymax = 7.5e-6) +
  theme_bw() +
  theme(aspect = 1, panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p2

pdf("p2.pdf", width = 4, height = 4)
p2
dev.off()
```

```{r fig.width=4, fig.height=4}
dt.ann.ab.s3 = dt.ann.ab %>%
  group_by(antigen.epitope, db.count.a, db.count.b) %>%
  summarise(count.ab = sum(count)) %>%
  ungroup %>%
  mutate(pab = count.ab / total / db.count.a / db.count.b) %>%
  select(antigen.epitope, pab) %>%
  merge(dt.ann.a.s %>%
          select(pa, antigen.epitope)) %>%
  merge(dt.ann.b.s %>%
          select(pb, antigen.epitope))

cres = with(dt.ann.ab.s3, cor.test(pa * pb, pab))
lbl = paste0("R=", round(cres$estimate, 2), ", P=", round(cres$p.value, 2))

p3=ggplot(dt.ann.ab.s3, aes(x = pa * pb, y = pab)) +
  annotate("text", x = 4e-10, y = 2e-9, label = lbl) +
  geom_smooth(method = "lm") +
  geom_text(aes(y = pab + 1e-10, label = substr(antigen.epitope, 1, 3))) +
  geom_point(size = 3) +
  xlab("(Mean TRA frequency) x (Mean TRB frequency)") +
  ylab("Mean TRA-TRB pair frequency") +
  theme_bw() +
  theme(aspect = 1, panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p3

pdf("p3.pdf", width = 4, height = 4)
p3
dev.off()
```

## Theoretical probabilities

```{r}
dt.vdjdb.pgen = fread("zcat vdjdb_ab_pgen.txt.gz") %>%
  select(cdr3.alpha, v.alpha, cdr3.beta, v.beta, antigen.epitope,
         genP_1mism_alpha, genP_1mism_beta) %>%
  unique %>%
  filter(antigen.epitope %in% dt.epi$antigen.epitope)
```

Summarise probabilities

```{r fig.width=4, fig.height=4}
dt.vdjdb.pgen.s = dt.vdjdb.pgen %>%
  group_by(antigen.epitope) %>%
  summarise(pa = mean(genP_1mism_alpha, na.rm=T),
            pb = mean(genP_1mism_beta, na.rm=T),
            pab = mean(genP_1mism_alpha * genP_1mism_beta, na.rm=T))

cres = with(dt.vdjdb.pgen.s, cor.test(pa, pb))
lbl = paste0("R=", round(cres$estimate, 2), ", P=", round(cres$p.value, 2))

p4=ggplot(dt.vdjdb.pgen.s, aes(x = pa, y = pb)) +
  annotate("text", x = 2e-5, y = 5e-6, label = lbl) +
  geom_smooth(method = "lm") +
  geom_point(size = 3) +
  geom_text(aes(y = pb + 3e-7, label = substr(antigen.epitope, 1, 3))) +
  xlab("Mean Pgen[TRA]") +
  ylab("Mean Pgen[TRB]") +
  theme_bw() +
  theme(aspect = 1, panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p4
pdf("p4.pdf", width = 4, height = 4)
p4
dev.off()

cres = with(dt.vdjdb.pgen.s, cor.test(pa * pb, pab))
lbl = paste0("R=", round(cres$estimate, 2), ", P=", round(cres$p.value, 3))

p5=ggplot(dt.vdjdb.pgen.s, aes(x = pa * pb, y = pab)) +
  annotate("text", x = 0.7e-10, y = 4.5e-10, label = lbl) +
  geom_smooth(method = "lm") +
  geom_text(aes(y = pab + 3e-11, label = substr(antigen.epitope, 1, 3))) +
  geom_point(size = 3) +
  xlab("(Mean Pgen[TRA]) x (Mean Pgen[TRB])") +
  ylab("Mean Pgen[TRA,TRB]") +
  theme_bw() +
  theme(aspect = 1, panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p5
pdf("p5.pdf", width = 4, height = 4)
p5
dev.off()

cres = with(dt.vdjdb.pgen.s, cor.test(pb, pab))
lbl = paste0("R=", round(cres$estimate, 2), ", P=", round(cres$p.value, 4))

p6=ggplot(dt.vdjdb.pgen.s, aes(x = pb, y = pab)) +
  annotate("text", x = 2e-6, y = 4.5e-10, label = lbl) +
  geom_smooth(method = "lm") +
  geom_text(aes(y = pab + 3e-11, label = substr(antigen.epitope, 1, 3))) +
  geom_point(size = 3) +
  xlab("Mean Pgen[TRB]") +
  ylab("Mean Pgen[TRA,TRB]") +
  theme_bw() +
  theme(aspect = 1, panel.grid.major = element_blank(), panel.grid.minor = element_blank())
p6
pdf("p6.pdf", width = 4, height = 4)
p6
dev.off()
```

